# Introduction to R {-}

The goal of this tutorial is to orient the learner to R Studio and the R programming language. 

Please complete the following:

1. Create a folder somewhere you can easily find it (e.g., on your desktop) called 'R Course'.
2. Open R Studio
3. Session -> Set Working Directory -> to source file location

## Outline {-}

1. Read in the 'babies' data set 
2. Basic data manipulation
3. Functions
4. Saving and viewing results

## Babies Dataset {-}

The  `babies` dataset  will be used throughout this session to illustrate basic R concepts. 

The dataset is a collection of variables taken for each new mother in a Child and Health Development Study.  It has  1,236 observations on male live births for the following 23 variables.

Variables in data file

```
id            identification number
date          birth date as character string (mon-dd-yyyy)
ddate         day of birth
mdate         month of birth
ydate         year of birth
gestation     length of gestation in days
wt            birth weight in ounces (999 unknown)
parity        total number of previous pregnancies 
              including fetal deaths and still births
age           mother's age in years at end of pregnancy
                999=unknown
ed            mother's education 
                 0=less than 8th grade
                 1=8th-12th grade - did not graduate
                 2=HS graduate???no other schooling
                 3=HS+trade
                 4=HS+some college
                 5=College graduate,
                 6=Trade school HS unclear
                 9=unknown
ht            mother's height in inches to the last completed inch 
                 999=unknown
wt1           mother prepregnancy wt in pounds
                 999=unknown
dage          father's age, coding same as mother's age.
ded           father's education, coding same as mother's education.
dht           father's height, coding same as for mother's height
dwt           father's weight coding same as for mother's weight
inc           family yearly income in $2500 increments
                 0=under 2500
                 1=2500-4999
                 ..., 
                 8= 12,500-14,999
                 9=15000+
                 998=unknown
                 999=not asked
smoke         does mother smoke? 
                 0=never
                 1=smokes now
                 2=until current pregnancy
                 3=once did, not now
                 9=unknown
time          If mother quit smoking, how long ago? 
                 0=never smoked
                 1=still smokes
                 2=during current preg
                 3=within 1 yr
                 4=1 to 2 years ago
                 5=2 to 3 years ago
                 6=3 to 4 years ago
                 7=5 to 9 years ago 
                 8=10+years ago
                 9=quit and don't know,
                 998=unknown
                 999=not asked
number        number of cigs smoked per day for past and current smokers 
                 0=never
                 1=1-4
                 2=5-9
                 3=10-14
                 4=15-19
                 5=20-29
                 6=30-39
                 7=40-60
                 8=60+
                 9=smoke but don't know
                 998=unknown
                 999=not asked
race          mother's race 
marital       marital status of mother
drace         father's race, coding same as mother's race
```

### Loading babies dataset {-}

We can can load the data from our working directory as shown below.  It is a .csv file, so can be read in with `read.csv`. specifying that 

- the file has a "header" row (`header=T`) with variable names 
- values are separated by commas.  
- values of '', '998' or '999' represent missing data

We need to be confident that 998 and 999 are not legitimate values for any variables, so that R does not interpret legitimate values as missing.  We will not distinguish between 'unknown' and 'not asked' in the data, so the code below will read two consecutive commas (``'), 998, and 999 as missing values.  We will have to deal separately with the values '9' that mean a value is missing.

```{r, include=FALSE}
knitr::opts_knit$set(root.dir = 'R/')
```

```{r}
babies <- read.csv('babies.csv',header=T, sep=",", 
                     na.strings=c("", "998", "999"))
```

### Data manipulation {-}

What are the dimensions of the data set?
```{r, results = 'asis', echo=T}
dim(babies)
```

What are the names of the variables?
```{r,  echo=T}
names(babies)
#head(babies)
#tail(babies)
```

Isolating variables
```{r, results = 'markup', echo=T}
# babies$id
# babies[, 1]
# babies[, 'id']
# babies[1:10, c('id', 'date', 'gestation')]
```

Assignment and subsetting
```{r, results = 'markup', echo=T}
under.30<- babies$age<30
b<- babies[under.30, ]
summary(b$age)
b<- babies[which(under.30), ]
summary(b$age)
```

#### Exercises (1) {-}

1. Create a new data set consisting of only the rows where the father's age is 40 or over (use '>=')

2. Include only the mother's age, father's age and marital status in the new dataset

3. Display the dimensions of this data set in the console.

4. Locate the new dataset in th Global Environment and view it

NB: Be sure that you have handled missing father's ages properly


##### Exercises (1) solutions {-}

```{r, eval=T, echo=params$SOLUTION,results=ifelse(!params$SOLUTION, "hide","asis")}
dad.over.40<- which(babies$dage>=40)
d<- babies[dad.over.40, c("age","dage","marital")]
dim(d)
# can also call the function directly
# View(d)
```

### Creating some new variables {-}

A baby weight more than 4kg is defined as macrosomia.  We can create this variable (converting ounces to kg) as:

```{r}
babies$wtKg <- babies$wt/(16*2.2)
babies$macrosomia <- babies$wtKg > 4
```

A baby weighing less than 2.5kg is defined as "small".  We can create this variable as:

```{r}

babies$smallBaby <- babies$wtKg < 2.5
```

We can make a single character variable with three categories of weight:

```{r}
babies$birthWeightCat <- ifelse(babies$wtKg < 2.5, "small",
                                ifelse(babies$wtKg > 4, "large","normal"))
```

Notice that when we tabulate these, they  appear in alphabetical order.  

```{r}
table(babies$birthWeightCat)
```

To impose an order, we can convert to  a factor and specify the order:

```{r}
babies$birthWeightCat <- factor(babies$birthWeightCat, 
                                  levels=c("small","normal","large"),
                                  labels=c("Small baby","Normal weight baby","Large baby"))
table(babies$birthWeightCat)                                
```

A baby born before 37 completed weeks of pregnancy is defined as preterm.  We can create this logical variable as:

```{r}
babies$preterm <- babies$gestation < 37*7
table(babies$preterm)
```

This is just a logical variable (`TRUE/FALSE` or `T/F`) so we need to make a factor to have it appear more user-friendly:

```{r}
babies$whenBorn <- factor(babies$preterm, 
                          levels=c(TRUE,FALSE), 
                          labels=c("Preterm","Full-term"))
```

We can make a binary variable for the mother's and father's races:

```{r}
babies$whiteRace <- ifelse(babies$race=="white","white","other")
babies$dwhiteRace <- ifelse(babies$drace=="white","white","other")
```

#### Exercises (2) {-}

1. Starting with the numeric variable `smoke`, in the babies dataset, create a new variable `smokeCat` in the data set that has three levels: `never smoker`, `past smoker`,`current smoker`.  Use the `factor` function and ensure that the levels appear in that order.
2. Tabulate the numerical variable against the new one to make sure you have not made an error. 
3. Make a binary variable in the babies dataset `smokeNow` that is 'Yes' when the mother smokes and 'No' otherwise (i.e., it is not yes (1) or it is  missing (9).)
4. [optional] Make a factor variable `eduCat` from the mother's education variable `ed` using the information at the top if this file about the meaning of 1,2,3,...

##### Exercises (2) solutions {-}

1. Make the two variables:

```{r, eval=T, echo=params$SOLUTION,results=ifelse(!params$SOLUTION, "hide","asis")}
babies$smokeCat <- with(babies, 
                        factor(smoke, 
                               levels=c(0,2,3,1),
                               labels=c("Never smoker","Past smoker","Past smoker","Current smoker")))

```

2. Check the coding was correct

```{r, eval=T, echo=params$SOLUTION,results=ifelse(!params$SOLUTION, "hide","asis")}
with(babies, table(smoke, smokeCat,useNA = 'always'), exclude=NULL)
```

3. Make a binary variable

```{r, eval=T, echo=params$SOLUTION,results=ifelse(!params$SOLUTION, "hide","asis")}
babies$smokeNow <- ifelse(babies$smoke==1 , "Yes", "No")
```

4. [optional] make the education variable and check it

```{r, eval=T, echo=params$SOLUTION,results=ifelse(!params$SOLUTION, "hide","asis")}
babies$eduCat <- factor(babies$ed, 
                        levels=0:6, 
                        labels=c("< 8th grade",
                                "8th-12th grade",
                                 "HS graduate",
                                  "HS+trade",
                                  "HS+some college",
                                  "College graduate",
                                  "Trade school/HS unclear"))
with(babies, table(ed, eduCat, useNA='always'))
```

## Functions {-}

Two types:

1. Built in functions
2. User defined functions

### Built in Functions {-}

When you open R, there are many functions available to you: Here, we will review a few useful built in functions. 

If you need help using a function, execute ? followed by the function name, with or without the parentheses.

```
? table
```

The following functions are widely used in descriptive statistics

#### table(), prop.table() {-}

```{r, echo=T}
t<- with(babies, table(marital))
t
p<- prop.table(t)
p
p*100
```

#### ifelse() {-}

```{r}
babies$first.preg<- with(babies, ifelse(parity==0, 'first','not first'))
table(babies$first.preg)

```

#### summary(), mean(), median(), sd(), quantile() {-}

```{r, echo=T}
summary(babies)
gestAge<- babies$gestation
mean(gestAge)
mean(gestAge, na.rm=T)

gestAge<- gestAge[complete.cases(gestAge)]
summary(gestAge)

mean(gestAge)
median(gestAge)
sd(gestAge)
range(gestAge)
min(gestAge)
max(gestAge)
quantile(gestAge)
quantile(gestAge, seq(0, 1, by = 0.2))
```

#### round() {-}

```{r, echo=T}

x<- 1121.933384
round(x, 3)
round(x, -2)
```

#### paste() {-}

We can build up complex quoted strings with `paste`:

```{r, echo=T}
paste("Mean (days) =", mean(gestAge))

paste("Mean (days) =", round(mean(gestAge)))


paste('mean day (sd) = ', 
      round(mean(gestAge)), 
      ' (', 
      round(sd(gestAge), 1), 
      ')', 
      sep='')

```

#### apply() {-}
 
Carries out an operation on the rows or columns of a dataset.  We pick the `MARGIN` to specify whether we want
this to be rows or columns

- 1 = rows
- 2 = columns

```{r, echo=T}

apply(babies[, c('age', 'dage', 'wtKg')], MARGIN = 2, FUN = mean, na.rm=T)

```

#### tapply() {-}

Carries out an operation on one variable, split by a second variable (or group of variables) - e.g., means by group:

```{r, echo=T}
wt.by.smoke.term<- with(babies, 
                       tapply(wt, list(smokeCat, whenBorn), mean, na.rm=T))
rw<- round(wt.by.smoke.term, 1)
rw
```

### User Defined Functions {-}

You can write your own functions, text(arguments){ body }

```{r, echo=T}
custom.summary<- function(x){
  out<- paste(round(mean(x, na.rm=T), 2), 
              ' (SD=', 
              round(sd(x, na.rm=T), 2), 
              ')', 
              sep='') 
  return(out)
}
custom.summary(babies$age)
```

### Saving output and opening in word {-}

```{r, echo=T}
write.csv(rw, 'test.csv', quote= F)

```

Now open 'test.csv' with word. Highlight and click Table -> convert text to table -> ok.

#### Exercises (3) {-}

1. Create a function to summarize a binary (0/1 or FALSE/TRUE)) variable and return a character string that looks like, for example "18/54 (33.3%)"

Framework

```
bin.sum <- function(x){
t <- # tabulate x
n <- # how many nonmissing observations?
x <- # how many 1's?
pct <- round()
paste()

}
```

2. Use `tapply` to apply your function to the macrosomia using smokeCat as the grouping variable

Hint

```
result <- tapply(X = 
                 INDEX= 
                 FUN=bin.sum)
        
```

3. View your results as a Word table using by using write.csv to save to a CSV file.

4. [Advanced] Create a function to summarize a continuous variable in the following format: "Median (IQR.low, IQR.high), n".  Use `apply` to apply this function to the variables `age`, `dage`, 'ht', 'dht'.  The result should be a table that you can save in the same way as in question 3.

5. [More advanced] Use the function from question 4. to summarize a the  variables `age`, `dage`, 'ht', 'dht' in groups formed by 'preterm'.  Hint: Look at the help  for the function `aggregate`.

Hint:

```
aggregate(x=DATASET, 
          by=list(BY VARIABLES HERE), 
          FUN=FUNCTION.FROM.4)
```


##### Exercises (3) solutions {-}

1. 
```{r, eval=T, echo=params$SOLUTION,results=ifelse(!params$SOLUTION, "hide","asis")}
bin.sum<- function(x){
  t<- table(x)
  pct<- round(100*prop.table(t), 1)
  out<- paste(t[2],
              "/",
              sum(t),
              " (", pct[2], "%)", sep="")
  out
}


x<- babies$macrosomia
bin.sum(babies$macrosomia)
```

2. 
```{r, eval=T, echo=params$SOLUTION,results=ifelse(!params$SOLUTION, "hide","asis")}
macro.smoke<- with(babies, tapply(macrosomia, smokeCat, bin.sum))
macro.smoke
```

3. 
```{r, eval=T, echo=params$SOLUTION,results=ifelse(!params$SOLUTION, "hide","asis")}
write.csv(macro.smoke, 'test2.csv', quote=F)

```

4. 
```{r, eval=T, echo=params$SOLUTION,results=ifelse(!params$SOLUTION, "hide","asis")}
custom.summary.q<- function(x){
  out<- paste(round(median(x, na.rm=T), 2), 
              ' (', 
              round(quantile(x, 0.25,na.rm=T), 2),
              ", ",
              round(quantile(x, 0.75,na.rm=T), 2), 
              ')', 
              sep='') 
  return(out)
}

apply(babies[,c("age", "dage", "ht", "dht")], MARGIN=2,FUN=custom.summary.q)
```



5. 
```{r, eval=T, echo=params$SOLUTION,results=ifelse(!params$SOLUTION, "hide","asis")}
aggregate(x=babies[,c("age","dage","ht","dht")], 
          by=list(babies$preterm), 
          FUN=custom.summary.q)
```

\newpage

## Working with more than one dataset {-}

Suppose you want to know how far from the average for the ethnic group each baby's birth weight is as a Z-score:

```
Z = (wt - average)/SD
```


You need to

1. find the average and SD for each group
2. merge this data with the full babies dataset
3. calculate each baby's Z score

The R code for these steps is shown below

1. find the average and SD for each group and save it in a data.frame

```{r}

average.by.group <- tapply(X = babies$wtKg, 
                           INDEX=babies$race, 
                           mean,na.rm=T)
                           
sd.by.group <- tapply(X = babies$wtKg, 
                     INDEX=babies$race, 
                     sd,na.rm=T)
stats <- data.frame(race=names(average.by.group),
                    average=average.by.group,
                    SD=sd.by.group)
print(stats)               
```

2. merge this data with the full babies dataset

```{r}
babies <- merge(babies, 
                stats, 
                by="race")

head(babies[ , c("wtKg","average", "SD","race")])
```



3. calculate each baby's Z score

```{r}
babies$Z.wt <- (babies$wtKg - babies$average)/babies$SD
```


Finally, please save the alterations to the babies data set for the next session:

```{r, echo=T}
write.csv(babies, 'babiesAugmented.csv', quote=F, row.names = F)
```


## Matrices {-}

A  data structure useful in statistical programming and for more advanced data analysis is a matrix.  A matrix is a 2-dimensional array, with rows and columns, that all contains items of the same type (as it is with vectors). We access their content using two indices (given the two dimensions).

```{r}
M1 <- matrix(1:12,nrow=3,byrow = F)	
M1a <- matrix(1:12,nrow=3,byrow = T)	
M2 <- matrix(letters[1:25],nrow=5,ncol = 5)	
M3 <- matrix(c(letters,LETTERS[1:4]),nrow=5,ncol = 6)	
```


```{r}
M3[1,1]	
M3[1:3,1:3]	
```
	
## Lists	{-}
	
The most versatile data structure is the `list`. It can be used as a collection of different, even heterogeneous, objects. It can have one or many dimentions (e.g. list array). We need to access a list's items using double square brackets `[[]]`.	

```{r}
some.stuff <- list(first.matrix=M1,	
                   second.matrix=M2, 	
                   my.data.frame=babies, 	
                   OldDads=babies$dage > 50)	
```

```{r}
class(some.stuff)	
length(some.stuff)	
names(some.stuff)	
some.stuff[[1]]	
table(some.stuff[["OldDads"]])	
	
some.stuff$my.vector[3]	
some.stuff$my.data.frame$gestation[41:45]
```
