--- 
title: "Advanced decision modelling in the context of Health Technology Assessment"
author: "Juan Pablo Diaz Martinez"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography:
- book.bib
- packages.bib
biblio-style: apalike
link-citations: yes
# description: This is a minimal example of using the bookdown package to write a book.
#   The output format for this example is bookdown::gitbook.
params:
  SOLUTION: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(dplyr)
library(readr)
```

# Introduction {-}

Bringing a new health technology to market and into the hands of a patient is a long process. Most of the times patients, who have a medical need, ask themselves why does it take so long to make the health technology available to everyone. When a health technology is in the market, it usually took between 5 to 10 years to make 
it available.

Depending on the country, governments usually are involved in the reimbursement process. They usually ask the next questions when a new health technology is available:

* How much does it cost?
* Will it save lives and/or improve quality of life?
* Do we have enough budget to fund it?
* If we have a pool of interventions for a specific disease, which one/ones should we reimburse?

Moreover, physicians, patients, insurance plans, and advocacy groups play an important role when new technologies are available in the market (why?). Even though a new technology see the light (i.e. it has proved to be safe and effective), insurance providers or the government will not necessarily cover it. Usually they argue that the new technology is "Not cost-effective" or "Not have good value for money". *These notes aim to provide all the necessary tools to decide if a new intervention has a good value-for-money.* It is important to stress that value-for-money decision is only one of many questions that are asked by one of the users of a **health technology assessment (HTA)**: patients, healthcare workers, government, and others.  

## Why reimbursment submissions fail? {-}

According to @goeree2015health, the reasons for rejection are:

1. Inappropriate comparator. Lack of proper statistical analysis.
2. Inappropriate outcome. Use of surrogates.
3. Inappropriate analysis. Lack of robust evidence for costs and quality of life.
4. High cost to the government. 

## Topics of the course {-}

1. What is HTA?
2. Introduction to decision-analytic models
3. Good practices in decision modelling
4. Evidence-based medicine
5. Decision tree-models
6. State-transition models with the Markov assumption
7. Partitioned survival models
8. Microsimulation
9. Discrete-event simulation
10. Uncertainty and decision-making
11. Presentation of results

## Statistical computing {-}

The use of open-source programming languages, such as `R`, in health decision sciences is growing and has the potential to facilitate model transparency, reproducibility, and shareability. However, realizing this potential can be challenging. Models are complex and primarily built to answer a research question, with model sharing and transparency relegated to being secondary goals. Moreover, many decision modelers are not formally trained in computer programming and may lack good coding practices, further compounding the problem of model transparency. **Therefore, throughout this course, the programming language `R` will be used to show its potential for advanced modelling in the context of HTA.**

For this course, we will be using the book ["R for Data Science"](https://r4ds.had.co.nz/). To install `R` and `Rstudio`, instructions are provided in Chapter 1 of this book. We will also use Excel throughout this course.

## Evaluation {-}

```{r table1, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- " 
| Item                | Percentage    | Due date      |
|:-------------------:|:-------------:|:-------------:|
| Assignment 1        | 15%           | Dec 5, 2021  |  
| Assignment 2        | 15%           | Dec 23, 2021  |  
| Take-home exam      | 30%           | Jan 7, 2022   |
| Project proposal    | 5%            | Dec 2, 2021  | 
| Project presentation| 5%            | Jan 14, 2022  | 
| Final project       | 30%           | Jan 17, 2022  | 
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

The intent is to  allow the students to demonstrate their mastery of this class through the following way. **Project proposal, presentation and final project will be done in pairs**.

### Asssignments {-}

The assignments are handed out approximately two weeks prior to the due date. Late work will not be marked, with the exception of an advance permission from the instructor.

### Project proposal {-}

(1 page)

The final deliverable for this course is a mini-HTA on a medical technology (preferably something  topical), with a focus on the quantitative aspect of it. Given that  the  translation of a health policy question into a relevant research question is an essential first step in the conduct of HTA, students are  required to formulate a research  question and submit for grading purposes. This should include at leas some of the following: an overview of the technology being assessed; a clear specification of the policy problem; and the research question(s) (including PICO) with objectives.

### Project presentation {-}

(20 minutes with extra 5 minutes for questions)

Students will be expected to present their final course paper and answer questions. Student will be graded on their presentations.

### Final project {-}

(20 pages double-spaced)

The main assignment will require students to produce a scaled down HTA, with a focus on the quantitative aspect of it. The objective of the final project is for the student to show that they have obtained a clear understanding of the advanced methods in decision modelling in the context of HTA. More information will be provided throughout the course, but the paper should contain the following:

a) Background and technology overview
b) Formulation of the question you are trying to answer through your mini-HTA
c) Review of the clinical literature
d) Description of the structure of the model
e) Description of the function of the model
f) Results
g) Conclusions

## Bibliography {-}

*Briggs, A., Sculpher, M., & Claxton, K. (2006). Decision modelling for health economic evaluation. Oxford University Press.*

*Gray, A. M., Clarke, P. M., Wolstenholme, J. L., & Wordsworth, S. (2011). Applied methods of cost-effectiveness analysis in healthcare (Vol. 3). Oxford University Press.*

*Edlin, R., McCabe, C., Hulme, C., Hall, P., & Wright, J. (2015). Cost effectiveness modelling for health technology assessment: a practical course. Springer.*





